---
title: "EAES 480 — Lab 03: Linear Models on an Andes-to-Amazon Gradient"
author: "Student: ____________________"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    df_print: paged
---

<!--
EAES 480 LAB CONTRACT
- Discuss concepts freely; type your own code; cite collaborators.
- Your work must knit (Run All) from a clean session.
- Put ALL code in chunks (not in the Console only).
-->

# Study context

This lab uses tidy site- and canopy-structure data derived from:

Asner, G.P., Anderson, C.B., Martin, R.E., Knapp, D.E., Tupayachi, R., Sinca, F., Malhi, Y. (2014)  
**Landscape-scale changes in forest structure and functional traits along an Andes-to-Amazon elevation gradient**. *Biogeosciences*, 11, 843–856. https://doi.org/10.5194/bg-11-843-2014

Your task is to build intuition for **simple linear models** (one predictor, one response), including:
- visualizing correlations,
- fitting regression lines,
- interpreting coefficients,
- and diagnosing fit using residuals.

---

# Learning goals

By the end of this lab you should be able to:

1. Join tidy datasets (sites + canopy structure) into one analysis table
2. Visualize bivariate relationships with scatterplots and fitted lines
3. Fit a simple linear model with `lm()`
4. Interpret coefficients using `broom::tidy()`
5. Overlay the fitted line with `geom_abline()` using the estimated coefficients
6. Assess model fit using residual plots (resid vs fitted, histogram/density, QQ plot)

---

# Deliverables + expectations

Submit:

- Your completed `.Rmd`
- A knitted HTML (or PDF if requested)

Rubric (lightweight):

- **Code correctness (40%)**: runs cleanly, outputs sensible
- **Interpretation quality (40%)**: answers reflect modeling + residual thinking
- **Reproducibility (20%)**: knits cleanly; no console-only objects

> Collaborators (if any): _______________________________

---

# Setup (run this first)

```{r setup, echo=TRUE, message=FALSE, warning=FALSE, eval = F}
# GOAL: Load packages, read data, join tables, and create a clean analysis dataset.

library(tidyverse)
library(janitor)
library(skimr)
library(broom) # used for looking at model fits

set.seed(___)  # TODO: choose an integer seed

# NOTE: In the course repo, put both CSVs in a ./data/ folder.
# TODO: Fill in file paths.
sites_path     <- "___"  # e.g., "data/andes_amazon_sites.csv"
structure_path <- "___"  # e.g., "data/andes_amazon_structure.csv"

sites_raw     <- readr::read_csv(sites_path)
structure_raw <- readr::read_csv(structure_path)

# Clean names to snake_case for easier typing
sites     <- sites_raw %>% clean_names()
structure <- structure_raw %>% clean_names()

# Join on site ID
df_raw <- left_join(structure, sites, by = "site")

# CHECKS
glimpse(df_raw)
skim(df_raw)
```

### Debugging checklist (before asking for help)

- Did you run the setup chunk first?
- Are your column names in **snake_case** (use `names(df_raw)`)?
- If you see `NA` values, did you filter them appropriately before modeling?
- Are your variables numeric (use `glimpse()`)?

---

# Data dictionary (quick)

Fill in at least **5** more rows, including your `x` and `y`. Reference the paper as needed.

```{r dictionary, echo=TRUE, eval = F}
names(df_raw)
```

| Column | Units | Meaning | Notes |
|---|---|---|---|
| site | id | site identifier | join key |
| elevation_m | m | elevation | predictor candidate |
| canopy_height_m | m | canopy height | response candidate |
| mat_c | °C | mean annual temperature | predictor candidate |
| map_mm | mm | mean annual precipitation | predictor candidate |
| soil_order | category | soil order | potential grouping |
| ___ |  |  |  |

**Prompt (2–3 sentences):** What does one row/observation represent here, and why do we need to *join* two tables?

> *Write your answer here.*

---

# Choose your model variables

You will fit a **simple linear regression**:

$$
y = \\\\beta_0 + \\\\beta_1 x + \\\\epsilon
$$

Choose:
- one numeric **response** `y_var`
- one numeric **predictor** `x_var`

Suggested pairs (pick ONE):
- `canopy_height_m` ~ `elevation_m`
- `gap_size_lambda` ~ `elevation_m`
- `canopy_shape_p_h` ~ `elevation_m`
- `canopy_height_m` ~ `mat_c`
- `canopy_height_m` ~ `map_mm`

```{r choose_model, echo=TRUE, eval = F}
y_var <- "___"  # TODO: response variable name
x_var <- "___"  # TODO: predictor variable name

y_var # use these as a reminder, for filling code below, use the actual variable name without ""
x_var
```

Create an analysis dataset with complete cases for your chosen variables:

```{r complete_cases, echo=TRUE, eval = F}
df <- df_raw %>%
  filter(
    !is.na(___), 
    !is.na(___)
  )

# CHECK
nrow(df)
glimpse(df)
```

**Prompt (1 sentence):** How has the filtering altered the dataset? Be specific.

> *Write your answer here.*

---

# Visualize the relationship (scatter + smooth)

## Scatterplot (no model yet)

```{r scatter, echo=TRUE, eval = F}
# GOAL: See the relationship before fitting a model.

ggplot(df, aes(x = ___, y = ___)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    x = ___,
    y = ,
    title = "Scatterplot"
  ) +
  theme_classic(base_size = 20)
```

**Prompt (2–3 sentences):** Describe the pattern. Is it roughly linear? Any obvious outliers or clusters?

> *Write your answer here.*

---

## Add a regression line with `geom_smooth(method = "lm")`

```{r smooth_lm, echo=TRUE, eval = F}
# GOAL: Add the best-fitting line (least squares) visually.

ggplot(df, aes(x = ___, y = ___)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = ___) +   # TODO: choose TRUE/FALSE
  labs(
    x = ___,
    y = ___,
    title = "Scatterplot with linear fit"
  ) +
  theme_classic(base_size = 20)
```

**Prompt (2 sentences):** What does the shaded band represent (if you kept `se = TRUE`)? Would you trust this model for prediction beyond the data range?

> *Write your answer here.*

---

# Fit a simple linear model with `lm()`

```{r fit_lm, echo=TRUE, eval = F}
# GOAL: Fit y ~ x using lm().

m1 <- lm(___ ~ ___, data = df)  # TODO: fill in y_var and x_var correctly
m1
```

## Coefficients with `tidy()`

```{r tidy_coef, echo=TRUE, eval = F}
coef_tbl <- tidy(m1)
coef_tbl
```

**Prompt (3 sentences):**
- Interpret the slope ($\\\\beta_1$) *with units*.
- Interpret the intercept ($\\\\beta_0$) and say whether it is physically meaningful here.
- Is the direction of the slope consistent with your scatterplot?

> *Write your answer here.*

---

# Plot the fitted line manually using `geom_abline()`

`geom_smooth(method="lm")` fits the model internally.  
Here you will extract the fitted coefficients and plot the line explicitly.

```{r abline, echo=TRUE, eval = F}
# GOAL: Overlay the fitted line using intercept + slope.
# TODO: Extract the intercept and slope from m1.

b0 <- coef(m1)[["___"]]     # TODO: intercept term name (usually "(Intercept)")
b1 <- coef(m1)[["___"]]     # TODO: slope term name (x variable)

b0
b1

ggplot(df, aes(x = ___, y = ___)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_abline(intercept = b0, slope = b1, linewidth = 1.2) +
  labs(
    x = ___,
    y = ___,
    title = "Same fitted line, drawn explicitly"
  ) +
  theme_classic(base_size = 20)
```

**Prompt (1–2 sentences):** Why is it useful to be able to draw the line from coefficients?

> *Write your answer here.*

---

# Model fit and residual diagnostics (DataCamp Ch. 3)

Residuals are where model assumptions live.

We’ll use `augment()` to get fitted values and residuals.

```{r augment, echo=TRUE, eval = F}
aug <- augment(m1)
glimpse(aug)
```

## Residuals vs fitted values

```{r resid_fitted, echo=TRUE, eval = F}
# GOAL: Look for patterns (curvature), changing variance (funnel), clusters.

ggplot(aug, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals vs fitted"
  ) +
  theme_classic(base_size = 20)
```

**Prompt (2–3 sentences):** Do you see curvature (nonlinearity), a funnel (heteroscedasticity), or clusters (missing grouping variable)?

> *Write your answer here.*

---

## Residual distribution (histogram + density)

```{r resid_hist, echo=TRUE, eval = F}
# GOAL: Check whether residuals are roughly symmetric and centered at 0.

ggplot(aug, aes(x = .resid)) +
  geom_histogram(aes(y = after_stat(density)), bins = ___, alpha = 0.5) +  # TODO bins
  geom_density(linewidth = 1.1) +
  labs(
    x = "Residuals",
    y = "Density",
    title = "Residual distribution"
  ) +
  theme_classic(base_size = 20)
```

**Prompt (2 sentences):** Are residuals symmetric? If not, what does that suggest about the model or the data?

> *Write your answer here.*

---

## QQ plot (normality check)

```{r qq, echo=TRUE, eval = F}
# GOAL: Compare standardized residuals to a Normal distribution.

ggplot(aug, aes(sample = .std.resid)) +
  stat_qq(alpha = 0.7) +
  stat_qq_line() +
  labs(
    x = "Theoretical quantiles",
    y = "Std. residual quantiles",
    title = "QQ plot of standardized residuals"
  ) +
  theme_classic(base_size = 20)
```

**Prompt (2 sentences):** Where does the QQ plot deviate most? What does that imply (tails/outliers/skew)?

> *Write your answer here.*

---

# Model summary (R² and RMSE)

```{r glance, echo=TRUE, eval = F}
glance(m1) %>% select(r.squared, adj.r.squared, sigma)
```

Interpretation:
- **R²**: fraction of variance explained (in-sample)
- **sigma**: typical residual size (RMSE) in units of **y**

**Prompt (2–3 sentences):** Does a high R² automatically mean a “good” environmental model? Explain with one example of what could still be wrong.

> *Write your answer here.*

---

# Interpretation checkpoints (graded writing)

### Checkpoint 1: Before the model (2–3 sentences)
Describe the relationship you expected between your chosen x and y. Why?

> *Write your answer here.*

### Checkpoint 2: After fitting (3–4 sentences)
Interpret the slope and evaluate whether the model seems plausible **mechanistically** (not just statistically).

> *Write your answer here.*

### Checkpoint 3: After residual checks (3–4 sentences)
Use your residual plots to diagnose the model. Name at least **one** assumption that looks reasonable and **one** that might be violated.

> *Write your answer here.*

### Checkpoint 4: Reporting (2–4 sentences)
If you could report **one figure** and **one table** from this lab to a environmental science reader, which would you choose and why?

> *Write your answer here.*

---

# 13. Extension 

Choose **one**.

## 13.1 Add a grouping variable (soil order) as color or facets

```{r extension_group, echo=TRUE, eval = F}
# GOAL: See whether grouping explains structure in the residuals.
# TODO: add color = soil_order (or facet_wrap).

ggplot(df, aes(x = ___, y = ___, color = ___)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic(base_size = 20) +
  labs(x = ___, y = ___, color = "soil_order")
```

**Prompt (2 sentences):** Does grouping change your interpretation of the relationship?

> *Write your answer here.*

## 13.2 Fit a second model and compare

Fit an alternative predictor and compare residual patterns and R².

```{r extension_compare, echo=TRUE, eval = F}
# TODO: choose an alternative predictor (e.g., mat_c instead of elevation_m)
m2 <- lm(___ ~ ___, data = df)

tibble(
  model = c("m1", "m2"),
  r2    = c(glance(m1)$r.squared, glance(m2)$r.squared),
  rmse  = c(glance(m1)$sigma,     glance(m2)$sigma)
)
```

**Prompt (2–3 sentences):** Which model would you trust more, and why? Use residual evidence.

> *Write your answer here.*

---

# 14. Submission + self-check

## Before you knit (Run All)
- [ ] Setup chunk runs without errors
- [ ] Data join happened (df_raw has both canopy + climate columns)
- [ ] All chunks run top-to-bottom
- [ ] No objects created only in the Console
- [ ] Interpretation answers are complete sentences

## After you knit
- [ ] Figures appear in the output
- [ ] Tables render and are readable
- [ ] Your selected x and y are clearly stated in the document

**Save, commit, push, and submit your GitHub repo link on Canvas.**
